---
title: "Mini-Project 01"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 01

For this report, I am using the dataset with U.S. births data. Each row contains 
the number of births each day starting from 2000 through 2014. There's a column
for year, month, date of month, date in the yyyy-mm-dd format, day of week, and
the number of births.

With these data, I want to tell a story about different US birth patterns across various
time scales and what causes them. There are many different types of events
that affect the number of babies born during a particular time period. For example, long
nationwide events may affect the yearly births, but may be harder to see looking at
the monthly or daily births. There are also smaller events that just affect
the specific day that a baby is born. For example, holidays can affect the 
particular day of the month that a baby is born; there tends to be fewer babies
born on popular holidays. The day of the week can have the same effect; more babies
tend to be born on weekdays compared to weekends. In addition, some months are more 
popular birth months than others.

To do this, I chose to examine three patterns. The first visualization shows the total yearly 
births from 2000 through 2014. The second visualization shows the average births 
for each day of the month. The third visualization shows the average births for each
day of the week across every month.

```{r}
library(tidyverse)
library(plotly)
```

```{r}
births <- read_csv("../data/us_births_00_14.csv", show_col_types = FALSE)
```

# Number of births per year

```{r}
yearly_births <- births |> 
  group_by(year) |> 
  summarize(births = sum(births)) |>
  mutate(tooltip = paste0(
    "<b>Year:</b> ", year, "<br>",
    "<b>Total Births:</b> ", scales::comma(births),
    ifelse(year == 2007, "<br><b>Note:</b> Baby boom peak", "")
  ))

p1 <- ggplot(yearly_births, aes(year, births)) +
  geom_line(color = "#1F77B4", group = 1, linewidth = 1) + 
  geom_point(aes(text = tooltip), color = "#1F77B4", size = 2.5) +
  # Highlight the 2007 Peak
  geom_vline(xintercept = 2007, linetype = "dashed", color = "#D32F2F", alpha = 0.7) +
  annotate("text", x = 2006.8, y = 4300000, label = "2007 Peak\n(4.32M)", hjust = 1, color = "#D32F2F", fontface = "bold", size = 3.5) +
  scale_x_continuous(breaks = 2000:2014, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M"), limits = c(3.8e6, 4.4e6)) +
  theme_minimal() +
  labs(title = "Number of births per year", 
       subtitle = "United States, 2000 to 2014",
       x = NULL, 
       y = "Number of births")

ggplotly(p1, tooltip = "text", height = 350, width = 800)
```

This is a line plot showing the relationship between the year on the x-axis, and
the number of births on the y-axis. We can see that between 2000 and 2014 there
was a baby boom peaking in 2007. There were over 200,000 more babies born in 
2007 compared to 2000. Starting in 2008, the number of births begins to fall off
linearly, reaching a low of about 4,000,000 in 2011, which holds stable through 2013. 
This low is over 100,000 less than in 2000. My first thought for the cause was the 
2008 financial crisis. I think this is a plausible cause considering the start of
the fall in births began the same year, and began to level off after the worst
of the effects of the financial crisis had happened.

For the graph, I showed a data point for each year, representing the number of
births, and connected the points with a line. The line shows the hump 
peaking in 2007 and the fall off immediately afterwards, demonstrating both
the baby boom and the financial crisis that followed.
For the design, I thought about Tufte's data-to-ink ratio and 
went with a simple grid and omitted the labels for the
x and y axes, which were redundant for this graph.

Originally, this graph used a trend line to show the "hump" in 2007. I replaced
it with lines to better represent the data, while still showing the "hump". I
thought about using a more granular x-axis to show this trend, because the data
has births for every day. I tried to show the births for every month, hoping it
would be as smooth as the yearly data, but with more points. Instead the data
was very messy and the "hump" was hardly visible.

# Average births by date-of-month

```{r}
dom <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
new_births <- births |>
  summarize(births = mean(births), .by = c(date_of_month, month)) |>
  complete(date_of_month = 1:31, month = 1:12) |>
  mutate(
    overall_avg = mean(births, na.rm = TRUE),
    month = factor(month, levels=1:12, labels = dom),
    pct_diff = (births - overall_avg) / overall_avg * 100,
    tooltip = ifelse(is.na(births), 
                     paste0("<b>", month, " ", date_of_month, "</b><br>Non-existent date"),
                     paste0(
                       "<b>", month, " ", date_of_month, "</b><br>",
                       "<b>Average births:</b> ", scales::comma(round(births)), "<br>",
                       "<b>Relative to Average:</b> ", 
                       ifelse(round(pct_diff, 1) == 0, "Equal to average", 
                              paste0(abs(round(pct_diff, 1)), "% ", 
                                     ifelse(pct_diff > 0, "above", "below"), " average"))
                     ))
  )
(new_births |>
  ggplot(aes(date_of_month, month, fill = births, text = tooltip)) +
  geom_tile(color="white") + 
  scale_x_continuous(breaks=1:31, minor_breaks = NULL) +
  scale_y_discrete(limits = rev) +
  scale_fill_viridis_c(labels=scales::label_comma(), na.value = "#EAEAEA") +
  theme_minimal() +
  labs(title = "Average births by day of month across every month",
       subtitle = "United States, 2000 to 2014",
       fill = "Average births",
       x = "Day of month",
       y = NULL)) |> ggplotly(tooltip = "text", height = 320, width = 800)
```
This graph shows the relationship for the average number of births for each day
of the month, across the full 15 year time span. 
We can see some relative lows on Jan 1st, Jul 4th, and Dec 25th, the last
week of Nov, Dec 31st, and Oct 31st. All of these are major holidays:
New Years, Independence Day, Thanksgiving, and Halloween. We can also see some
relative highs during the summer months, especially September. One particular
day that stands out to me is Apr 1st. It is noticeably darker than the surrounding
days, but there is no holiday then. I'm not sure what would have caused it, it's possibly
related to April Fools' Day but it seems unlikely to me.

For the graph, I used a 2d heatmap plot. The result clearly shows which days of 
the month have relatively more or fewer births compared to their neighbors and
correlates well with recognizable holidays, which have a preattentive pop-out 
that makes them immediately obvious. I chose the viridis theme specifically
because it is a sequential color scheme that makes it very easy to see slight 
variations in births compared to other themes. I also applied the data-to-ink
ratio principle and removed the axis labels, because they are redundant.

This visualization went through some revisions. Initially, I had a bar chart, with
a bar for every month, with all the days of each month averaged together. That
graph wasn't very interesting and the small changes were hard to gauge.
Then, I made a chart that
averaged together every month and placed the days in a grid made to resemble
a calendar. I decided to instead use this graph, which shows the holidays
more clearly. Also, the calendar view might be a misrepresentation because it
does not show days of the week.

# Average births by day-of-week and month
```{r}
dow <- c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")
heatmap_data <- births |>
  mutate(day_of_week = factor(day_of_week, levels=dow),
         month = factor(month, levels= 1:12, labels=dom)) |>
  summarize(births = mean(births), .by=c(day_of_week, month)) |>
  mutate(tooltip = paste0(
    "<b>Month:</b> ", month, "<br>",
    "<b>Day of Week:</b> ", day_of_week, "<br>",
    "<b>Average births:</b> ", scales::comma(round(births))
  ))

p3_heatmap <- heatmap_data |>
  ggplot(aes(day_of_week, month, fill = births, text = tooltip)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(labels = scales::label_comma()) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(title = "Average births by day-of-week and month",
       subtitle = "United States, 2000 to 2014",
       x = NULL,
       y = NULL,
       fill = "Average births")

ggplotly(p3_heatmap, tooltip = "text", height = 320, width = 800)
```
This graph shows the relationship between the day of the week and the average number
of births. We can see that there are relatively fewer births on the weekends
compared to the weekdays across every month. We can also see that September has
a higher number of births compared to the rest of the months. A particularly
interesting point is Thursdays in November; there is a small drop compared to
the surrounding cells. This is probably Thanksgiving, because it is always on
Thursday.

Like the previous graphs, this is a 2d heatmap. I used gestalt rules to place Saturday
and Sunday next to each other, which highlights their similarity of relatively low
births. I believe it makes it more obvious that the cause is the weekend. 
I also used the sequential viridis color scheme, like the previous graph.
Also, I applied the data-to-ink ratio principle and removed the axis labels, like the other
two graphs.

This graph did not have any revisions. I did experiment with placing the weekend
data in different places, on each end and at the beginning.

```{r}
line_data <- births |>
  mutate(day_of_week = factor(day_of_week, levels=dow),
         month = factor(month, levels=1:12, labels=dom)) |>
  summarize(births = mean(births), .by=c(day_of_week, month)) |>
  mutate(dow_avg = mean(births), .by=c(day_of_week)) |>
  mutate(monthy_avg = mean(births), .by=c(month)) |>
  mutate(
    avg_births = mean(births),
    pct_diff = (births - avg_births) / avg_births * 100,
    pct_diff_dow = (births - dow_avg) / dow_avg * 100,
    pct_diff_monthly = (births - monthy_avg) / monthy_avg * 100,
    tooltip = paste0(
    "<b>Month:</b> ", month, "<br>",
    "<b>Day of Week:</b> ", day_of_week, "<br>",
    "<b>Average births:</b> ", scales::comma(round(births)), "<br>",
    "<b>Relative to ", day_of_week,":</b> ", 
     ifelse(round(pct_diff_dow, 1) == 0, "Equal to average", 
       paste0(abs(round(pct_diff_dow, 1)), "% ", 
      ifelse(pct_diff_dow > 0, "above", "below"), " average")), "<br>",
     "<b>Relative to ", month,":</b> ", 
     ifelse(round(pct_diff_monthly, 1) == 0, "Equal to average", 
       paste0(abs(round(pct_diff_monthly, 1)), "% ", 
      ifelse(pct_diff_monthly > 0, "above", "below"), " average"))
  ))

p3_line <- line_data |>
  ggplot(aes(x = month, y = births, group = day_of_week, color = day_of_week)) +
  geom_line(linewidth = 1) +
  geom_point(aes(text = tooltip), size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(title = "Average births by day-of-week and month",
       subtitle = "Grouped Line Chart (United States, 2000 to 2014)",
       x = "Month",
       y = "Average births",
       color = "Day of week")

ggplotly(p3_line, tooltip = "text", height = 400, width = 800)
```

These alternative charts show the same relationship but highlight different patterns. By shifting from a 2D heatmap to grouped bar and line charts, the dramatic drop on weekends (Saturday and Sunday) compared to weekdays is much more visually distinct.

In the grouped line chart particularly, we can see two distinct bands: a lower band of two lines representing weekends (Sat and Sun, hovering around 8,000 births) and an upper band of five lines representing weekdays (hovering around 12,000 births). This visual separation immediately highlights the weekend drop, which is much clearer than comparing colors on a heatmap grid.

In addition, the seasonal trend (with births peaking in late summer/September) is visible across all days of the week, with all lines/bars shifting upward during those months. We can also see a drop in births on Thursdays in November (the line for Thursday dips in November), which is due to Thanksgiving always falling on a Thursday.
