---
title: "Mini-Project 01"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 01

For this report, I am using the dataset with U.S. births data. Each row contains 
the number of births each day starting from 2000 through 2014. There's a column
for year, month, date of month, date in the yyyy-mm-dd format, day of week, and
the number of births.

With these data, I want to tell a story about different US birth patterns across various
time scales and what causes them. There are many different types of events
that affect the number of babies born during a particular time period. For example, long
nationwide events may affect the yearly births, but may be harder to see looking at
the monthly or daily births. There are also smaller events that just affect
the specific day that a baby is born. For example, holidays can affect the 
particular day of the month that a baby is born; there tends to be fewer babies
born on popular holidays. The day of the week can have the same effect; more babies
tend to be born on weekdays compared to weekends. In addition, some months are more 
popular birth months than others.

To do this, I chose to examine three patterns. The first visualization shows the total yearly 
births from 2000 through 2014. The second visualization shows the average births 
for each day of the month. The third visualization shows the average births for each
day of the week across every month.

```{r, message = FALSE}
library(tidyverse)
library(plotly)
```

```{r}
births <- read_csv("../data/us_births_00_14.csv", show_col_types = FALSE)
```

# Number of births per year

```{r, warning=FALSE, fig.alt = 'Line chart showing the total number of births in the United States'}
yearly_births <- births |> 
  group_by(year) |> 
  summarize(births = sum(births)) |>
  mutate(tooltip = paste0(
    "<b>Year:</b> ", year, "<br>",
    "<b>Total Births:</b> ", scales::comma(births)
  ))

(ggplot(yearly_births, aes(year, births)) +
  geom_line(color = "#1F77B4", group = 1, linewidth = 1) + 
  geom_point(aes(text = tooltip), color = "#1F77B4", size = 2.5) +
  geom_vline(xintercept = 2007, linetype = "dashed", color = "#D32F2F", alpha = 0.7) +
  annotate("text", x = 2007.7, y = 4400000, label = "2007 Peak\n(4.38M)", hjust = 1, color = "#D32F2F", fontface = "bold", size = 3.5) +
  scale_x_continuous(breaks = 2000:2014, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M"), limits = c(3.8e6, 4.4e6)) +
  theme_minimal() +
  labs(title = "Number of births in the United States", 
       subtitle = "United States, 2000 to 2014",
       x = NULL, 
       y = "Number of U.S. births")) |> ggplotly(tooltip = "text")
```

This is a line plot showing the relationship between the year on the x-axis, and
the number of births on the y-axis. We can see that between 2000 and 2014 there
was a baby boom peaking in 2007. There were over 200,000 more babies born in 
2007 compared to 2000. Starting in 2008, the number of births begins to fall off
linearly, reaching a low of about 4,000,000 in 2011, which holds stable through 2013. 
This low is over 100,000 less than in 2000. My first thought for the cause was the 
2008 financial crisis. I think this is a plausible but unconfirmed cause considering the start of
the fall in births began the same year, and began to level off after the worst
of the effects of the financial crisis had happened.

For the final revision of this graph, I added interactivity. Now when hovering
over a data point, the exact number of births for that year will be displayed.

I made some small improvements as well including adding a vertical line showing 
the peak number of births in 2007, increasing the thickness of the lines, abbreviating
the number of births, and adding an axis label. With these changes, the chart
looks better, is more self-contained, and shows the point of interest of this chart more clearly.

For comparison, this is the original graph:

```{r echo=FALSE, fig.alt='Line chart showing the total number of births per year in the United States from 2000 to 2014 (original layout)'}
births |> 
  group_by(year) |> 
  summarize(births = sum(births)) |> 
  ggplot(aes(year, births)) +
  geom_point() +
  geom_line() + 
  scale_x_continuous(breaks = 2000:2014, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::label_number()) +
  theme_minimal() +
  labs(title = "Number of births per year", 
       subtitle = "United States, 2000 to 2014",
       x = NULL, 
       y = NULL)
```

# Average births by date-of-month
```{r, fig.alt = 'Heatmap showing the average number of births in the United States from 2000 to 2014 across each date of the month for each month.'}
dom <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
new_births <- births |>
  summarize(births = mean(births), .by = c(date_of_month, month)) |>
  complete(date_of_month = 1:31, month = 1:12) |>
  mutate(
    overall_avg = mean(births, na.rm = TRUE),
    month = factor(month, levels=1:12, labels = dom),
    pct_diff = (births - overall_avg) / overall_avg * 100,
    tooltip = ifelse(is.na(births), 
                     paste0(""),
                     paste0(
                       "<b>", month, " ", date_of_month, "</b>",
                       ifelse(month == "Dec" & date_of_month == 25, ", <b>Christmas Day</b><br>", 
                              ifelse(month == "Jul" & date_of_month == 4, ", <b>Independence Day</b><br>",
                                     ifelse(month == 'Jan' & date_of_month == 1, ", <b>New Years' Day</b><br>",
                                            ifelse(month == 'Oct' & date_of_month == 31, ", <b>Halloween</b><br>",
                                                   ifelse(month == 'Nov' & date_of_month >= 22 & date_of_month <= 28, ", <b>Thanksgiving week</b><br>",
                                                          ifelse(month == 'Dec' & date_of_month == 31, ", <b>New Years' Eve</b><br>",
                                                                 ifelse(month == 'Dec' & date_of_month == 24, ", <b>Christmas Eve</b><br>",
                              "<br>"))))))),
                       "<b>Average births:</b> ", scales::comma(round(births)), "<br>",
                       "<b>Relative to Average:</b> ", 
                       ifelse(round(pct_diff, 1) == 0, "Equal to average", 
                              paste0(abs(round(pct_diff, 1)), "% ", 
                                     ifelse(pct_diff > 0, "above", "below"), " average"))
                     ))
  )
(new_births |>
  ggplot(aes(date_of_month, month, fill = births, text = tooltip)) +
  geom_tile(color="white") + 
  scale_x_continuous(breaks=1:31, minor_breaks = NULL) +
  scale_y_discrete(limits = rev) +
  scale_fill_viridis_c(labels=scales::label_comma(), na.value = "#EAEAEA") +
  theme_minimal() +
  labs(title = "Average births in the United States by day of month across every month",
       fill = "Average U.S. births",
       x = "Day of month",
       y = NULL)) |> ggplotly(tooltip = "text")
```
This graph shows the relationship for the average number of births for each day
of the month, across the full 15 year time span. 
We can see some relative lows on Jan 1st, Jul 4th, and Dec 25th, the last
week of Nov, Dec 31st, and Oct 31st. All of these are major holidays:
New Years, Independence Day, Thanksgiving, and Halloween. We can also see some
relative highs during the summer months, especially September. One particular
day that stands out to me is Apr 1st. It is noticeably darker than the surrounding
days, but there is no holiday then. I'm not sure what would have caused it, it's possibly
related to April Fools' Day but it seems unlikely to me.

For the final revision of this graph, I added interactivity. Hovering over
a cell in the heatmap shows the month, day of month, average number of births (floored),
a major holiday if one exists, and a percent error of that cell's birth count with the overall average.
The percent error is accessible substitute for the color of each cell. This also has the advantage
of showing precisely how many more births there are on that day of month compared
to the overall average.

For improvements, I added a label for the x-axis to make this graph more self-contained.
Visually, the graph looks the same as the original, so I will omit a comparison.

# Average births by day-of-week and month
```{r, warning=FALSE, fig.alt = 'A set of line charts showing the average number of births in the United States from 2000 through 2013 across every month with one line chart per day of the week. '}
dow <- c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")
line_data <- births |>
  mutate(day_of_week = factor(day_of_week, levels=dow),
         month = factor(month, levels=1:12, labels=dom)) |>
  summarize(births = mean(births), .by=c(day_of_week, month)) |>
  mutate(dow_avg = mean(births), .by=c(day_of_week)) |>
  mutate(monthy_avg = mean(births), .by=c(month)) |>
  mutate(
    avg_births = mean(births),
    pct_diff = (births - avg_births) / avg_births * 100,
    pct_diff_dow = (births - dow_avg) / dow_avg * 100,
    pct_diff_monthly = (births - monthy_avg) / monthy_avg * 100,
    tooltip = paste0(
    "<b>Month:</b> ", month, "<br>",
    "<b>Day of Week:</b> ", day_of_week, "<br>",
    "<b>Average births:</b> ", scales::comma(floor(births)), "<br>",
    "<b>Relative to ", day_of_week,":</b> ", 
     ifelse(round(pct_diff_dow, 1) == 0, "Equal to average", 
       paste0(abs(round(pct_diff_dow, 1)), "% ", 
      ifelse(pct_diff_dow > 0, "above", "below"), " average")), "<br>",
     "<b>Relative to ", month,":</b> ", 
     ifelse(round(pct_diff_monthly, 1) == 0, "Equal to average", 
       paste0(abs(round(pct_diff_monthly, 1)), "% ", 
      ifelse(pct_diff_monthly > 0, "above", "below"), " average"))
  ))

(line_data |>
  ggplot(aes(x = month, y = births, group = day_of_week, color = day_of_week)) +
  geom_line(linewidth = 1) +
  geom_point(aes(text = tooltip), size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(title = "Average births in the U.S. from 2000 to 2014 by day-of-week and month",
       subtitle = "Grouped Line Chart (United States, 2000 to 2014)",
       x = NULL,
       y = "Average U.S. births",
       color = NULL)) |> ggplotly(tooltip = "text")
```

This graph shows the relationship between the day of the week and the average number
of births. We can see that there are relatively fewer births on the weekends
compared to the weekdays across every month. We can also see that September has
a higher number of births compared to the rest of the months. A particularly
interesting point is Thursdays in November; there is a small drop compared to
the surrounding cells. This is probably Thanksgiving, because it is always on
Thursday.

For the final revision, I changed this graph from a 2D heatmap to a set of line charts.
With the line charts, the distinct drop in births during the weekends is still visible,
but is not reliant on color to convey this fact. In addition, the line charts show
trends across the months for each day of the week more obviously than a 2D heatmap.
For example, with the line charts, we can see a seasonal trend where births peak
in the summer, which can be seen every day of the week, especially the weekend. This
increase can be seen on the 2D heatmap, but the trend is less obvious.

I also added interactivity, so hovering over each data point will display the (floored) average births,
and a percent error relative to both the day of the week and the month. This extra information
makes changes quantifiable. For example, by looking at Thursdays in November, we can
see that the average number of births is 9.3% below average compared to Thursdays in other months but 4.6% above average other days of the week in November.

For comparison, this is the original graph:

```{r echo=FALSE, fig.alt='A heatmap showing the average number of births in the United States from 2000 to 2014 across each day of the week for each month.'}
heatmap_data <- births |>
  mutate(day_of_week = factor(day_of_week, levels=dow),
         month = factor(month, levels= 1:12, labels=dom)) |>
  summarize(births = mean(births), .by=c(day_of_week, month)) |>
  mutate(tooltip = paste0(
    "<b>Month:</b> ", month, "<br>",
    "<b>Day of Week:</b> ", day_of_week, "<br>",
    "<b>Average births:</b> ", scales::comma(round(births))
  ))

heatmap_data |>
  ggplot(aes(day_of_week, month, fill = births, text = tooltip)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(labels = scales::label_comma()) +
  scale_y_discrete(limits = rev) +
  theme_minimal() +
  labs(title = "Average births by day-of-week and month",
       subtitle = "United States, 2000 to 2014",
       x = NULL,
       y = NULL,
       fill = "Average births")
```
