---
title: "Mini-Project 02"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 02

For this project, I am using the FIFA18 dataset. It is sourced from the FIFA18 EA Sports video game.
Each row contains information about a player including name, nationality, club, age, and 34 attributes
that quantify the player's performance in the game on a scale of 0 to 100.

With these data, I want to look at comparisons between the overall player rating and
the various other attributes to show how they affect the overall rating in the game. The overall
rating is correlated with all 34 attributes including other columns such as nationality, which I
will demonstrate with three visualizations. The first is a map that displays average overall score for each country.
The second is a scatter plot that fits a linear model to the 34 attributes and compares the result to the overall score.
The third is a box-and-whisker plot that shows which attributes best predict overall rating.

```{r, message=FALSE}
library(plotly)
library(tidyverse)
library(sf)
library(htmlwidgets)
library(htmltools)
library(ggrepel)
library(ggtext)
library(broom)
```

```{r}
fifa <- read_csv("../data/fifa18.csv", show_col_types = FALSE)
world_shapes <- read_sf("../data/ne_110m_admin_0_countries/ne_110m_admin_0_countries.shp")|>
    filter(ISO_A3 != "ATA")
```

# 1: Average Overall Rating by Country
```{r, warning=FALSE}
fifa_map <- world_shapes |>
  left_join(
    fifa |> summarize(n = n(), avg_score = mean(overall), .by = c(nationality)),
    by = join_by(NAME_LONG == nationality)
  ) |>
  mutate(
    n = replace_na(n, 0),
    hover = paste0(
      "<b>", NAME_LONG, "</b><br>",
      "Total Players: ", n, "<br>",
      "Average Rating: ", if_else(n > 0, as.character(round(avg_score, 1)), "N/A")
    )
  )

fifa_map |>
  plot_geo() |>
  add_trace(z = ~avg_score, 
            text = ~hover, 
            color = ~avg_score, 
            colors = 'Purples', 
            locations = ~ISO_A3,
            hoverinfo = "text") |>
  colorbar(title = "Average Player Rating") |>
  layout(title = "Average FIFA18 player rating by country",
         geo = list(showframe = TRUE, 
                    showcoastlines = TRUE, 
                    projection = list(type = 'robinson')))
```

This is an interactive map showing the average overall player rating for each country.
It shows countries with more players with a dark purple and countries with no
players as white. When hovering over a country it shows the precise average overall score,
the country name, and the number of players from that country.

We can see that there are some clusters of high performing areas. In South America
Brazil, Peru, Venezuela, Paraguay, and Ecuador have a rating of roughly 70. There
is a small cluster around Northern Africa that includes Algeria, Morocco, Spain,
and Portugal with a rating of roughly 70 as well. The same pattern can be seen in 
Eastern Europe with Belarus, Ukraine, Czech Republic, Serbia, Bosnia and Herzegovina, and Croatia.
Other countries such as the United States, the United Kingdom, Germany, Poland, Sweden,
Finland, Canada, Mexico, Russia, China, and Australia have lower scores of about 65 or less.
Now, some countries such as the United Kingdom have many players (1920) but has a an
average score of 63, while other countries such as Belarus has few players (9) but an
average score of 70. It's possible in some countries it is difficult to become a FIFA-level soccer
player, so there might be a survivorship bias for countries like Belarus. 
In the case of the United Kingdom, it's possible that because there are so many, the average decreases.

For the final revision, I switched from the mercator projection to the robinson projection, which
I believe looks nicer. 

# 2: Predicted vs. Actual Overall Rating
```{r, warning=FALSE}
all_attrs <- c(
  "acceleration", "aggression", "agility", "balance", "ball_control",
  "composure", "crossing", "curve", "dribbling", "finishing",
  "free_kick_accuracy", "gk_diving", "gk_handling", "gk_kicking",
  "gk_positioning", "gk_reflexes", "heading_accuracy", "interceptions",
  "jumping", "long_passing", "long_shots", "marking", "penalties",
  "positioning", "reactions", "short_passing", "shot_power",
  "sliding_tackle", "sprint_speed", "stamina", "standing_tackle",
  "strength", "vision", "volleys"
)

model_data <- fifa |>
  select(name, nationality, overall, all_of(all_attrs)) |>
  drop_na()

fit <- lm(overall ~ ., data = model_data |> select(-name, -nationality))

model_data <- model_data |>
  mutate(
    predicted = fitted(fit),
    residual  = residuals(fit)
  )

to_label <- bind_rows(
  model_data |> slice_max(residual, n = 4),
  model_data |> slice_min(residual, n = 4)
)

tooltip_text <- function(name, nationality, predicted, overall, residual) {
  paste0(
    "<b>", name, "</b><br>",
    nationality, '<br>',
    "Predicted: ", round(predicted, 1), "<br>",
    "Actual: ", overall, "<br>",
    "Residual: ", round(residual, 2), "<br>",
    ifelse(predicted > overall, "Overrated", "Underrated")
  )
}

model_data <- model_data |>
  mutate(tooltip = tooltip_text(name, nationality, predicted, overall, residual))

to_label <- to_label |>
  mutate(tooltip = tooltip_text(name, nationality, predicted, overall, residual))

(ggplot(model_data, aes(x = predicted, y = overall)) +
  geom_point(aes(color = residual, text = tooltip), alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(data = to_label, aes(color = residual, text = tooltip), size = 3) +
  geom_text(data = to_label, aes(label = name), size = 2.5, vjust = -0.8) +
  scale_color_gradient2(
    low = "steelblue", mid = "grey85", high = "tomato",
    midpoint = 0, name = "Residual"
  ) +
  labs(
    title = "Predicted vs. Actual Overall Rating for players in FIFA18",
    x = "Predicted Rating",
    y = "Actual Rating"
  ) +
  theme_minimal() +
  theme(legend.position = "none")) |> ggplotly(tooltip = "text")
```

This interactive scatter plot compares a predicted overall rating vs the actual overall rating.
To calculate the predicted overall rating, I fit a linear model using the actual overall
with the other 34 attributes. It shows players where the actual overall rating was greater than
the predicted in red, and the opposite in blue. The players with a predicted value less than
the actual value are marked as "underrated" and "overrated" for the opposite condition. The
plot shows labels for the top 8 most overrated and underrated players. When hovering, there is
a tooltip with the player's name, nationality, predicted value, actual value, residual value, along with the underrated/overrated label.

We can see that the predicted rating correlates well with the true rating. Near the top
end of the range, the prediction tends to underrate, leaving almost no overrated players once
the actual rating reach roughly 85. There are some surprising outliers. Some deviate from the actual rating by
over 10 points. This is unexpected, I thought originally that there would be a much tighter spread. I'm not
sure what methods were used to calculate the overall value. I added in the countries to the tooltip because I was
curious if there would be any correlation. Just from looking at the outliers, there does not seem to be anything obvious.

I made no changes for the final revision.

# 3: Coefficient Plot
```{r}
coef_data <- tidy(fit, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    clean_term = str_replace_all(term, "_", " ") |> str_to_title(),
    category = case_when(
      str_detect(term, "^gk_") ~ "Goalkeeping",
      term %in% c("acceleration", "agility", "balance", "reactions", "sprint_speed", "stamina", "jumping", "strength") ~ "Physical / Athletic",
      term %in% c("aggression", "interceptions", "marking", "sliding_tackle", "standing_tackle", "heading_accuracy") ~ "Defending",
      term %in% c("ball_control", "crossing", "curve", "long_passing", "short_passing", "vision") ~ "Passing & Control",
      TRUE ~ "Attacking & Shooting"
    ),
    tooltip = paste0(
      "<b>", clean_term, "</b> (", category, ")<br>",
      "Correlates with ", ifelse(estimate > 0, "better", "worse"), " performance<br>",
      "Estimate: ", round(estimate, 3), "<br>",
      "p-value: ", format.pval(p.value, digits = 2), ifelse(p.value < 0.05, " (High Confidence)", " (Low Confidence)")
    )
  ) |>
  arrange(category, estimate) |>
  mutate(term = fct_inorder(clean_term))

plot_ly(
  coef_data,
  type = "scatter",
  mode = "markers",
  height = 900,
  x = ~estimate,
  y = ~term,
  color = ~category,
  colors = c(
    "Goalkeeping" = "#1f77b4",
    "Physical / Athletic" = "#ff7f0e",
    "Defending" = "#2ca02c",
    "Passing & Control" = "#9467bd",
    "Attacking & Shooting" = "#d62728"
  ),
  text = ~tooltip,
  hoverinfo = "text",
  error_x = list(
    type = "data",
    symmetric = FALSE,
    array = ~(conf.high - estimate),
    arrayminus = ~(estimate - conf.low),
    color = "gray60",
    thickness = 1.5,
    width = 4
  ),
  marker = list(size = 8)
) |>
  layout(
    title = "Which attributes best predict Overall Rating for players in FIFA18?",
    xaxis = list(title = "Coefficient Estimate", zeroline = TRUE),
    yaxis = list(title = "", dtick = 1),
    shapes = list(list(
      type = "line", x0 = 0, x1 = 0, y0 = 0, y1 = 1,
      yref = "paper", line = list(color = "grey", dash = "dash")
    )),
    legend = list(title = list(text = "Attribute Category"))
  )
```

This visualization shows the correlation between the player's attributes and the player's overall rating.
Positive coefficients mean the attribute is correlated with better overall performance, 
and a negative attribute means the coefficient is correlated with worse overall performance.
Hovering over a datapoint shows the category, effect on overall rating, coefficient estimate, and p-value.

Looking at the attributes by themselves, we can see that the attribute that best predicts a good overall rating is "Reactions" with a value of 0.291
and the attribute that best predicts a bad overall rating is "positioning" with a value of -0.054.
Most of the attributes are correlated positively with overall score. There were only 7 that were negatively correlated, compared to the 27 positively correlated attributes.
The "reactions" attribute is a large outlier that correlates almost 2 times as much as the next most positive attribute, "Ball Control".
Now, It makes sense that these are positive, because those attributes are two important skills that are needed in soccer.
I'm unsure why there are any attributes that negatively correlate, even a little bit. For example, the "positioning" attribute has
a near zero p-value, but why would positioning by negatively correlated?

Looking at the attribute categories, we can see a general "S-pattern" for each. 
There tends to be outliers for both relatively low and high coefficient values with a several more 
having a similar effect on rating for each category, creating an "S" shape.

For the final revision, I made several changes. First, I created categories to
group each attribute by, colored each datapoint based on category, and sorted each datapoint
by category, then estimate. The previous graph colored the datapoints by the sign of the coefficient.
I think it is more useful to color the datapoints by category because the positive/negative correlation is still clear with the vertical line.
With this information, we can see how the category of each attribute correlates with overall rating,
however, the categories are arbitrary and might not represent a perfect grouping.
I also changed the height of the graph to accommodate every attribute, so now there are no hidden attributes
revealed by zooming in or scrolling. 
For the last changes, I tweaked the tooltip information by omitting the confidence interval, added a line
that plainly states how the attributes correlates with rating, and added a confidence note for the p-value.

Here is the original graph for comparison:

```{r, echo=FALSE}
coef_data <- tidy(fit, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term      = str_replace_all(term, "_", " ") |> str_to_title(),
    direction = if_else(estimate > 0, "Positive", "Negative"),
    tooltip   = paste0(
      "<b>", term, "</b><br>",
      "Estimate: ", round(estimate, 3), "<br>",
      "95% CI: [", round(conf.low, 3), ", ", round(conf.high, 3), "]<br>",
      "p-value: ", signif(p.value, 3)
    )
  ) |>
  arrange(estimate) |>
  mutate(term = fct_inorder(term))

plot_ly(
  coef_data,
  type = "scatter",
  mode = "markers",
  x = ~estimate,
  y = ~term,
  color = ~direction,
  colors = c("Negative" = "steelblue", "Positive" = "tomato"),
  text = ~tooltip,
  hoverinfo = "text",
  error_x = list(
    type = "data",
    symmetric = FALSE,
    array = ~(conf.high - estimate),
    arrayminus = ~(estimate - conf.low),
    color = "gray60",
    thickness = 1.5,
    width = 4
  ),
  marker = list(size = 8)
) |>
  layout(
    title = "Which attributes best predict Overall Rating for players in FIFA18?",
    xaxis = list(title = "Coefficient Estimate", zeroline = TRUE),
    yaxis = list(title = ""),
    shapes = list(list(
      type = "line", x0 = 0, x1 = 0, y0 = 0, y1 = 1,
      yref = "paper", line = list(color = "grey", dash = "dash")
    )),
    legend = list(title = list(text = "Direction"))
  )
```